---
title: "Basic GIS plotting"
output: html_document
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)

# Download "Admin 0 – Countries" from
# https://www.naturalearthdata.com/downloads/110m-cultural-vectors/
world_map <- read_sf("data/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp")

# Download cb_2018_us_state_20m.zip under "States" from
# https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html
us_states <- read_sf("data/cb_2018_us_state_20m/cb_2018_us_state_20m.shp")

# Download cb_2018_us_county_5m.zip under "County" from
# https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html
us_counties <- read_sf("data/cb_2018_us_county_5m/cb_2018_us_county_5m.shp")

# Download "Admin 1 – States, Provinces" from
# https://www.naturalearthdata.com/downloads/10m-cultural-vectors/
us_states_hires <- read_sf("data/ne_10m_admin_1_states_provinces/ne_10m_admin_1_states_provinces.shp")
```

## World map

### Clean data

Because this is just a data frame, we can do all our normal dplyr things to it. Let's get rid of Antarctica, since it takes up a big proportion of the southern hemisphere:

```{r}
world_sans_antarctica <- world_map %>% 
  filter(ISO_A3 != "ATA")
```

Ready to plot a map? Here's all you need to do:

```{r}
ggplot() + 
  geom_sf(data = world_sans_antarctica)
```

### Your turn

Use the fill, color, and size options to change the country fill color, the border color, and the border width

INSERT CHUNK HERE


### More dynamic fill colors

The Natural Earth dataset happens to come with some columns with a coloring scheme with 7–13 colors (`MAPCOLOR7`, `MAPCOLOR9`, etc.) so that no countries with a shared border share a color. We can fill by that column:

```{r}
ggplot() + 
  geom_sf(data = world_sans_antarctica, 
          aes(fill = as.factor(MAPCOLOR7)),
          color = "#401D16", size = 0.25) +
  scale_fill_viridis_d(option = "plasma") +  # Use a nicer palette
  guides(fill = "none") +  # Remove the legend
  theme_void()  # Get rid of all the gridlines and extra visual clutter
```


### Projections

Changing map projections is trivial: add a `coord_sf()` layer where you specify the CRS you want to use. Here's Robinson (I found the ESRI code at [epsg.io](http://epsg.io/54030)):

```{r}
ggplot() + 
  geom_sf(data = world_sans_antarctica, 
          fill = "darkblue", color = "white", size = 0.25) +
  coord_sf(crs = st_crs("ESRI:54030")) +  # Robinson
  theme_void()
```

### Your turn

Make three different world maps using different projections (like Mercator, Equidistant cylindrical, Sinusoidal, Mollweide, etc. experiment and have fun!), fills, colors, sizes, etc.

INSERT CHUNKS HERE


## US map

### Basic map

This same process works for any shapefile. The map of the US can also be projected differently—two common projections are NAD83 and Albers. We'll take the `us_states` dataset, remove Alaska, Hawaii, and Puerto Rico (they're so far from the rest of the lower 48 states that they make an unusable map), and plot it.

```{r}
lower_48 <- us_states %>% 
  filter(!(NAME %in% c("Alaska", "Hawaii", "Puerto Rico")))

ggplot() + 
  geom_sf(data = lower_48, fill = "darkred", color = "white", size = 0.25) +
  coord_sf(crs = st_crs("EPSG:4269")) +  # NAD83
  theme_void()
```

### Your turn

Plot the 48 states using an Albers projection.

INSERT CHUNK HERE


### US map with non-contiguous parts

Plotting places like Alaska, Hawaii, and Puerto Rico gets a little tricky since they're far away from the contiguous 48 states. There's an easy way to handle it though!

First, there's a package named [**tigris**](https://github.com/walkerke/tigris) that provides a neat interface for working with spatial data from the [US Census's TIGER shapefiles](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html) (like downloading them directly for you). **tigris** is on CRAN, but as of May 2021, it's an older version from July 2020 that's missing some neat additions. Install the latest development version [following the instructions at GitHub](https://github.com/walkerke/tigris) (i.e. run `remotes::install_github('walkerke/tigris')` in your console).

In addition to providing a ton of functions for getting shapefiles for states, counties, voting districts, Tribal areas, military bases, and dozens of other things, **tigris** has a `shift_geometry()` function that will change the coordinates for Alaska, Hawaii, and Puerto Rico so that they end up in Mexico and the Gulf of Mexico.

```{r message=FALSE}
library(tigris)

# This is the Census shapefile we loaded earlier. Note how we're not filtering
# out AK, HI, and PR now
us_states_shifted <- us_states %>% 
  shift_geometry()  # Move AK, HI, and PR around

ggplot() +
  geom_sf(data = us_states_shifted) +
  coord_sf(crs = st_crs("ESRI:102003")) +  # Albers
  theme_void()
```

`shift_geometry()` has an optional `position` argument that lets you control where the non-contiguous states go. By default they'll go below the continental US (`position = "below"`), but you can also use `position = "outside"` to place them more in relation to where they are in real life.

### Your turn

Make a US map that includes Alaska, Hawaii, and Puerto Rico that uses different colors, fills, border sizes, and places AK, HI, and PR outside the lower 48 rather than below.

INSERT CHUNK HERE


## Single state maps

Again, because these shapefiles are really just fancy data frames, we can filter them with normal dplyr functions. Let's plot just Georgia:

```{r georgia-only}
only_georgia <- lower_48 %>% 
  filter(NAME == "Georgia")

ggplot() +
  geom_sf(data = only_georgia, fill = "purple") +
  theme_void()
```

There's one small final issue though: we're missing all the Atlantic islands in the southeast like Cumberland Island and Amelia Island. That's because we're using the Census's low resolution (20m) data. That's fine for the map of the whole country, but if we're looking at a single state, we probably want better detail in the borders. We can use the Census's high resolution (500k) data, but even then it doesn't include the islands for whatever reason, but [Natural Earth has high resolution US state data](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-admin-1-states-provinces/) that *does* have the islands, so we can use that:

```{r}
only_georgia_high <- us_states_hires %>% 
  filter(iso_3166_2 == "US-GA")

ggplot() +
  geom_sf(data = only_georgia_high, fill = "purple") +
  theme_void()
```


### Your turn

Make a map of some other state (not Georgia):

INSERT CHUNK HERE

