---
title: "Introduction"
author: "Andrew Heiss"
date: "July 13, 2021"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css"]
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view"))
xaringanExtra::use_share_again()
```

class: center middle section-title section-title-1 animated fadeIn

# Introduction to<br>GIS data with R

---

class: middle section-title-inv-1

.box-1.large[Why do *you* want to<br>learn about maps and GIS?]

.box-1.large[Where/when do you use<br>maps and GIS in your job?]

---

layout: true
class: title title-1

---

# Plan for today

.box-inv-1[**Maps, truth, and data**<br>.small[(30 minutes)]]

--

.box-inv-1[**Intro/overview of R, ggplot, and dplyr**<br>.small[(60 minutes)]]

--

.box-inv-1[**GIS in R with SF**<br>.small[(75 minutes)]]

---

# Resources

.box-inv-1[All workshop materials<br>and a ton of additional resources<br>are available here:]

.box-1.large[https://andhs.co/gpl-gis]

```{r echo=FALSE, include=FALSE, eval=FALSE}
pagedown::chrome_print(here::here("slides", "00_introduction.html"), 
                       output = here::here("slides", "00_introduction.pdf"))
```
